### 2024-07-09 paper/tex/05-proof.tex
edit and prepare for jfp

\begin{theorem*}[Soundness of Typing]
    \label{thm:soundness-typing}
    Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
    \hfill
    \begin{itemize}
        \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$,
        \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$,
        \item [$\bullet$] $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$, 
            for any instantiation of $[[uσ]]$ satisfying constraints $[[SC]]$.
            All of it under assumptions that $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢ uN]]$ and that $[[uN]]$ is free from 
            \emph{negative} algorithmic variables.
    \end{itemize}
\end{theorem*}

@@

The proof of soundness and completeness result is done gradually
for all the subroutines,
following the structure of the algorithm 
(\cref{fig:alg-typing-graph,fig:alg-subtyping-graph})
bottom-up. Next, we discuss the main of these results.

\subsection{Normalization}
    The point of type normalization is factoring out non-trivial equivalence 
    by selecting a representative from each equivalence class.
    This way, \emph{the correctness of normalization} means that
    checking for equivalence of two types is the same as checking for equality of their normal forms.
    \begin{lemma*}[Normalization correctness]
        Assuming all types are well-formed in $[[Γ]]$, we have
            $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ and 
            $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$.
    \end{lemma*}
    To prove the correctness of normalization, 
    we introduce an \emph{intermediate} relation on types---\emph{declarative equivalence}
    (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
    but explicitly allows quantifier reordering and redundant quantifier removal.
    Then the statement $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ (as well as its positive counterpart) 
    is split into two steps: $[[Γ ⊢ iN ≈ iM]] \iff [[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.

@@

The initiality shows that the anti-unifier that the algorithm provides is the
most specific (or the most `detailed'). Specifically, it means that any other
sound anti-unification solution can be `refined' to the result of the algorithm.  
The `refinement' is represented as an instantiation of the anti-unifier---a 
substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$ replacing the `holes' from $[[Ξ1]]$ with
types that themselves can contain `holes' from $[[Ξ2]]$.

@@

    The Least Upper Bound algorithm finds the least type that is a supertype of two given types.
    Its \emph{soundness} means that the returned type is indeed a supertype of the given ones;
    the \emph{completeness} means that the algorithm succeeds if the least upper bound exists;
    and the \emph{initiality} means that the returned type is the least among common supertypes.

@@

Algorithmic typing relies on the \emph{minimal instantiation} procedure.
For a given positive type $[[uP]]$ and a set of constraints $[[SC]]$
it returns a substitution $[[Θ ⊢ uσ : SC]]$ such that
$[[ [uσ]uP ]]$ is a subtype of all the other instantiations of $[[uP]]$ satisfying $[[SC]]$.
This way, the correctness of minimal instantiation is formulated as the following two lemmas.

\begin{lemma*}[Soundness of Minimal Instantiation]
    \label{lemma:minimal-instantiation-soundness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
    Then `$[[Γ ⊢ uP SC minby uσ ]]$' implies that 
    $[[Θ ⊢ uσ : uv uP]]$ is a normalized substitution satisfying $[[SC]]$
    (\ie $[[Θ ⊢ uσ : SC]]$)
    and for any other substitution $[[Θ ⊢ uσ' : uv uP ]]$ satisfying $[[SC]]$,
    we have $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.
\end{lemma*}

\begin{lemma*}[Completeness of Minimal Instantiation]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, $[[Γ; dom(Θ) ⊢ uP]]$, and there exists 
    $[[Θ ⊢ uσ : uv uP ]]$ satisfying $[[SC]]$ ($[[Θ ⊢ uσ : SC]]$) 
    such that for any other $[[Θ ⊢ uσ' : uv uP ]]$
    satisfying $[[SC]]$, we have
    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.  Then the minimal instantiation procedure  
    succeeds; specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ holds.
\end{lemma*}

The soundness is rather straightforwardly proved by induction on the 
judgment inference tree, relying on the soundness of the singularity procedure.
The proof of completeness is done by induction on the structure of the type $[[uP]]$.
It uses the fact that $[[ Γ ⊢ ∃nas.[uσ']uP ≥ ∃nas.[uσ]uP]]$ implies
$[[ Γ, nas ⊢ [uσ']uP ≥ [uσ]uP]]$.

@@

Finally, we discuss the proofs of the soundness and completeness of type
inference algorithm that we stated at the beginning of this section.
There are three subtleties that we will cover that are important 
for the proof to go through: the determinacy of the algorithm,
the mutual dependence of the soundness and completeness proofs, and
the non-trivial inductive metric that we use.

\paragraph*{Determinacy}
    One of the properties that our proof relies on is the determinacy of the 
    typing algorithm: the output (the inferred type)
    is uniquely determined by the input (the term and the contexts).
    Determinacy is not hard to demonstrate by structural induction: 
    in every algorithmic inference system, only one inference rule can be applied 
    for a given input.  However, we need to prove it for \emph{every} 
    subroutine of the algorithm. Ultimately, it requires 
    the determinacy of such procedures as \emph{generation of fresh variables}, 
    which is easy to ensure, but must be taken into account in the implementation.

\begin{lemma*}[Determinacy of the Typing Algorithm]
    Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ]]$. Then 
    \begin{itemize}
        \item [$+$] If $[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ v : iP']]$ then $[[iP = iP']]$.
        \item [$-$] If $[[Γ; Φ ⊨ c : iN]]$ and $[[Γ; Φ ⊨ c : iN']]$ then $[[iN = iN']]$.
        \item If $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ and 
            $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC']]$ then 
            $[[uM = uM']]$, $[[Θ]] = [[Θ']]$, and $[[SC]] =[[SC']]$.  
    \end{itemize}
\end{lemma*}

@@

Typically in our inductive proofs, the soundness is proven before completeness,
as the completeness requires the premise subtrees to satisfy certain properties
(which can be given by the soundness). However, in the case of the typing algorithm,
the soundness and completeness proofs cannot be separated: the inductive proof
of one requires another, and vice versa.

@@

    The size of \emph{declarative} and \emph{algorithmic} judgments is defined as
    a pair: the first component is the syntactic size of the terms used in the judgment. 
    The second component depends on the kind of judgment. For regular
    type inference judgments (such as $[[Γ ; Φ ⊢ v : iP]]$
    or $[[Γ ; Φ ⊨ c : iN]]$), it is always zero. For application inference judgments
    $([[Γ ; Φ ⊢ iN ● args ⇒> iM]]$ or $[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$), 
    it is equal to the number of prenex quantifiers of the head type $[[iN]]$.
    We need this adjustment to ensure the monotonicity of the metric, since the rules 
    \ruleref{\ottdruleATForallAppLabel} and \ruleref{\ottdruleDTForallAppLabel}
    only reduce the quantifiers in the head type but they do not change the list of arguments.

@@

This metric is suitable for mutual induction on the soundness and completeness
of the typing algorithm.  First, it is monotonous \wrt the inference rules,
and this way, we can always apply the induction hypothesis to premises 
of each rule. Second, the induction hypothesis is powerful enough,
so we can use the completeness of the algorithm in the soundness proof,
where required. For instance, to prove the soundness
of typing in case of $[[Γ; Φ ⊨ let x = v(args); c' : iN]]$,
we can assume, that  \emph{completeness}
holds for a term of shape $[[Γ; Φ ⊢ iM ● args ⇒> iK ]]$, since
$\size{args} < \size{let x = v(args); c'}$.
This is exactly what allows us to 
deal with the case of \ruleref{\ottdruleATAppLetLabel},
because then we can conclude that 
the inferred type (of a declarative judgment $[[Γ; Φ ⊢ iM ● args ⇒> ↑[uσ]uQ
]]$ constructed by the induction hypothesis) is unique.

>>>>

\begin{theorem*}[Soundness of Typing]
    \label{thm:soundness-typing}
    Suppose that $[[Γ ⊢ Φ]]$. Then\footnotemark[1]
    \begin{itemize}
        \item [$+$] $[[Γ; Φ ⊨ v : iP]]$ implies $[[Γ; Φ ⊢ v : iP]]$,
        \item [$-$] $[[Γ; Φ ⊨ c : iN]]$ implies $[[Γ; Φ ⊢ c : iN]]$,
        \item [$\bullet$] $[[Γ; Φ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$ implies $[[ Γ ; Φ ⊢ [uσ]uN ● args ⇒> [uσ]uM ]]$, 
            for any instantiation of $[[uσ]]$ satisfying constraints $[[SC]]$.
            All of it under assumptions that $[[Γ ⊢ Θ]]$ and $[[Γ; dom(Θ) ⊢ uN]]$ and that $[[uN]]$ is free from 
            \emph{negative} algorithmic variables.
    \end{itemize}
\end{theorem*}

@@

The proof of soundness and completeness result is done gradually
for all the subroutines,
following the structure of the algorithm 
(\cref{fig:alg-typing-graph,fig:alg-subtyping-graph})
bottom-up. Next, we discuss the main results.

\subsection{Normalization}
    The point of type normalization is factoring out non-trivial equivalence 
    by selecting a representative from each equivalence class.
    This way, \emph{the correctness of normalization} means that
    checking for the equivalence of two types is the same as checking for the equality of their normal forms.
    \begin{lemma*}[Normalization correctness]
        Assuming all types are well-formed in $[[Γ]]$, we have
            $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ and 
            $[[Γ ⊢ iP ≈ iQ]] \iff [[nf(iP) = nf(iQ)]]$.
    \end{lemma*}
    To prove the correctness of normalization, 
    we introduce an \emph{intermediate} relation on types---\emph{declarative equivalence}
    (the notation is $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$).
    In contrast to $[[Γ ⊢ iN ≈ iM]]$ (which means mutual subtyping), 
    $[[iN ≈ iM]]$ does not depend on subtyping judgments, 
    but explicitly allows quantifier reordering and redundant quantifier removal.
    Then the statement $[[Γ ⊢ iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$ (as well as its positive counterpart) 
    is split into two steps: $[[Γ ⊢ iN ≈ iM]] \iff [[iN ≈ iM]] \iff [[nf(iN) = nf(iM)]]$.

@@

The initiality shows that the anti-unifier that the algorithm provides is the
most specific (or the most `detailed'). Specifically, it means that any other
sound anti-unification solution can be `refined' to the result of the algorithm.  
The `refinement' is represented as an instantiation of the anti-unifier---a 
substitution $[[Γ ; Ξ2 ⊢ ausr : Ξ1]]$---replacing the `holes' from $[[Ξ1]]$ with
types that themselves can contain `holes' from $[[Ξ2]]$.

@@

    The Least Upper Bound algorithm finds the least type that is a supertype of two given types.
    Its \emph{soundness} means that the returned type is indeed a supertype of the given ones;
    the \emph{completeness} means that the algorithm succeeds if the least upper bound exists;
    and the \emph{initiality} means that the returned type is the least among the common supertypes.

@@

Algorithmic typing relies on the \emph{minimal instantiation} procedure.
For a given positive type $[[uP]]$ and a set of constraints $[[SC]]$
it returns a substitution $[[Θ ⊢ uσ : SC]]$ such that
$[[ [uσ]uP ]]$ is a subtype of all the other instantiations of $[[uP]]$ satisfying $[[SC]]$.
The correctness of minimal instantiation is formulated as the following two lemmas.

\begin{lemma*}[Soundness of Minimal Instantiation]
    \label{lemma:minimal-instantiation-soundness}
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, and $[[Γ; dom(Θ) ⊢ uP]]$.
    Then the judgment `$[[Γ ⊢ uP SC minby uσ ]]$' implies that 
    $[[Θ ⊢ uσ : uv uP]]$ is a normalized substitution satisfying $[[SC]]$
    (\ie $[[Θ ⊢ uσ : SC]]$)
    and for any other substitution $[[Θ ⊢ uσ' : uv uP ]]$ satisfying $[[SC]]$,
    we have $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.
\end{lemma*}

\begin{lemma*}[Completeness of Minimal Instantiation]
    Suppose that $[[Γ ⊢ Θ]]$, $[[Θ ⊢ SC]]$, $[[Γ; dom(Θ) ⊢ uP]]$, and there exists 
    $[[Θ ⊢ uσ : uv uP ]]$ satisfying $[[SC]]$ ($[[Θ ⊢ uσ : SC]]$) 
    such that for any other $[[Θ ⊢ uσ' : uv uP ]]$
    satisfying $[[SC]]$, we have
    $[[Γ ⊢ [uσ']uP ≥ [uσ]uP ]]$.  Then the minimal instantiation procedure  
    succeeds; specifically, $[[Γ ⊢ uP SC minby nf(uσ) ]]$ is derivable.
\end{lemma*}

Soundness is rather straightforwardly proved by induction on the 
judgment inference tree, relying on the soundness of the singularity procedure.
The proof of completeness is done by induction on the structure of the type $[[uP]]$.
It uses the fact that $[[ Γ ⊢ ∃nas.[uσ']uP ≥ ∃nas.[uσ]uP]]$ implies
$[[ Γ, nas ⊢ [uσ']uP ≥ [uσ]uP]]$.

@@

Finally, we discuss the proofs of the soundness and completeness of the type
inference algorithm that we described at the beginning of this section.
There are three subtleties that are important 
for the proof to go through: the determinacy of the algorithm,
the mutual dependence of the soundness and completeness proofs, and
the non-trivial inductive metric that we use.

\vspace{\baselineskip}
\paragraph*{Determinacy}
    One of the properties that our proof relies on is the determinacy of the 
    typing algorithm: the output (the inferred type)
    is uniquely determined by the input (the term and the contexts).
    Determinacy is not hard to demonstrate by structural induction: 
    in every algorithmic inference system, only one inference rule can be applied 
    for a given input.  However, we need to prove it for \emph{every} 
    subroutine of the algorithm. Ultimately, it requires 
    the determinacy of such procedures as \emph{generation of fresh variables}, 
    which is easy to ensure, but must be taken into account in the implementation.

\begin{lemma*}[Determinacy of the Typing Algorithm]
    Suppose that $[[Γ ⊢ Φ]]$ and $[[Γ ⊢ Θ0]]$. Then 
    \begin{itemize}
        \item [$+$] If $[[Γ; Φ ⊨ v : iP]]$ and $[[Γ; Φ ⊨ v : iP']]$ then $[[iP = iP']]$.
        \item [$-$] If $[[Γ; Φ ⊨ c : iN]]$ and $[[Γ; Φ ⊨ c : iN']]$ then $[[iN = iN']]$.
        \item If $[[Γ; Φ; Θ0 ⊨ uN ● args ⇒> uM ⫤ Θ; SC]]$ and 
            $[[Γ; Φ; Θ0 ⊨ uN ● args ⇒> uM' ⫤ Θ'; SC']]$ then 
            $[[uM = uM']]$, $[[Θ]] = [[Θ']]$, and $[[SC]] =[[SC']]$.  
    \end{itemize}
\end{lemma*}

@@

Typically in our inductive proofs, the soundness is proven before completeness,
as the completeness requires the premise subtrees to satisfy certain properties
(which can be given by the soundness). However, in the case of the typing algorithm,
the soundness and completeness proofs cannot be separated: the inductive proof
of one requires the other.

@@

    The size of \emph{declarative} and \emph{algorithmic} judgments is defined
    as a pair: the first component is the cumulative syntactic size of the terms
    used in the judgment. The second component depends on the kind of judgment.
    For regular type inference judgments (such as $[[Γ ; Φ ⊢ v : iP]]$ or 
    $[[Γ ; Φ ⊨ c : iN]]$), it is always zero. For application inference judgments $([[Γ
    ; Φ ⊢ iN ● args ⇒> iM]]$ or $[[Γ ; Φ ; Θ ⊨ uN ● args ⇒> uM ⫤ Θ'; SC]]$), it
    is equal to the number of prenex quantifiers of the head type $[[iN]]$. We
    need this adjustment to ensure the monotonicity of the metric, since the
    quantifier instantiation rules \ruleref{\ottdruleATForallAppLabel} and
    \ruleref{\ottdruleDTForallAppLabel} only reduce the quantifiers in the head
    type but they do not change the list of arguments.

@@

This metric is suitable for mutual induction on the soundness and completeness
of the typing algorithm.  First, it is monotone \wrt the inference rules,
and this way, we can always apply the induction hypothesis to premises 
of each rule. Second, the induction hypothesis is powerful enough,
so we can use the completeness of the algorithm in the soundness proof,
where required. For instance, to prove the soundness
of typing in case of $[[Γ; Φ ⊨ let x = v(args); c' : iN]]$,
we can assume, that  \emph{completeness}
holds for a term of shape $[[Γ; Φ ⊢ iM ● args ⇒> iK ]]$, since
$\size{args} < \size{let x = v(args); c'}$.
This is exactly what allows us to 
deal with the case of \ruleref{\ottdruleATAppLetLabel},
because then we can conclude that 
the inferred type (of a declarative judgment $[[Γ; Φ ⊢ iM ● args ⇒> ↑[uσ]uQ
]]$ constructed by the induction hypothesis) is unique.

### 2024-07-09 paper/tex/06-extensions.tex
edit and prepare for jfp

In our system, all type applications are inferred implicitly: the algorithm
automatically instantiates the variables abstracted by $\forall$. 
The \emph{explicit} type application can be added to the declarative system by the following rule:
$$\ottdruleDTTypeAppLabeled{}$$

@@

\paragraph*{Solution 1: Declarative Normalization}
One way to resolve this ambiguity is to fix the order of quantifiers.  
The algorithm already performs the ordering of the quantifiers in the normalization procedure.
This way, we could require the inferred type to be normalized
to specify the order of the quantifiers:
$$\ottdruleDTTypeAppOrderedLabeled{}$$

@@

\paragraph*{Solution 2: Elementary Type Inference}
An alternative approach to provide explicit type application
 was proposed by \cite{zhao22:elementary}.
In this work, the subtyping relation is restricted 
in such a way that 
$[[∀α⁺.∀β⁺.iN]]$ and $[[∀β⁺.∀α⁺.iN]]$ are \emph{not} mutual subtypes 
(as long as $[[α⁺ ∊ fv(iN)]]$ and $[[β⁺ ∊ fv(iN)]]$).
It implies that the order of the quantifiers of the inferred type is unique,
and thus, the explicit type application is unambiguous.

@@

The elementary type inference restricts the expressiveness of the subtyping perhaps too much.
As mentioned, it forbids the quantifier reordering; besides, 
as soon as the right-hand side of the subtyping relation is
polymorphic, it restricts the instantiation of the left-hand side quantifiers, 
which \emph{disallows}, for example, $[[· ⊢ ∀α⁺.↑α⁺ ≤ ∀α⁺.↑↓↑α⁺]]$.
On the other hand, the elementary subtyping system
can be smoothly extended with bounded quantification, which we discuss later.

\paragraph*{Solution 3: labeled quantifiers}

A compromise solution that resolves the ambiguity of explicit instantiation
without fixing the order of quantifiers is by using \emph{labeled} quantifiers.
Let us discuss how to introduce them to the \emph{negative} part of the system,
since the positive subsystem is symmetric.

@@

The subtyping rule permits an arbitrary order of the quantifiers.
However, similarly to the elementary subtyping, 
each of the right-hand side quantifiers must have 
a matching left-hand side quantifier with the same label. These
matching quantifiers synchronously removed
(in other words, each variable is abstractly instantiated to itself), 
and the remaining quantifiers are instantiated as usual by
a substitution $[[σ]]$.
$$ \ottdruleDOneLblForallLRLabeled{} $$

@@

Second, notice that the positive subtyping is still not dependent on the negative
subtyping. It means, that any leaf-to-root path in the inference tree is monotonous:
in the first phase, the negative subtyping rules are applied, and in the second phase, 
only the positive ones.

@@

We do not present the details of the constraint resolution procedure or how it
is integrated into the subtyping algorithm, leaving it for future work. However,
we believe that the system consisting of the mentioned kinds of constraints is solvable
by the following reasons.
\begin{enumerate*}
    \item[(i)] The equivalence entries $[[pua :≈ uP]]$ and $[[nua :≈ uN]]$ are resolved by standard first-order pattern unification techniques.
    \item[(ii)] The resolution of supertyping entries ($[[pua :≥ iP]]$) is discussed in the original algorithm (\cref{sec:constraint-merge}).
    \item[(iii)] The new kind of entries $[[pua :≤ uP]]$ can be
        reduced by case analysis. If $[[uP]]$ is a variable $[[β⁺]]$
        or a shifted computation $[[↓uN]]$ then 
        $[[pua :≤ uP]]$ is equivalent to $[[pua :≈ uP]]$;
        otherwise, $[[pua :≤ ∃nas.↓uN]]$ is equivalently replaced by $[[pua :≈ [nuas/nas]↓uN]]$
        for freshly created $[[nuas]]$.
\end{enumerate*}

@@

\paragraph*{The Algorithm}

    The algorithmic system is bidirectionalized in a similar way. 
    Each typing premise and the conclusion of the rule is oriented
    to either `checking` ($[[⇐]]$) or `synthesizing` ($[[⇒]]$) mode,
    with respect to the declarative system.

@@

    % \paragraph*{Restricting Inference}
    Instead, we strengthen
    \ruleref{\ottdruleATArrowAppBidirLabel}
    in such a way that it never checks a term against an algorithmic type.
    Instead, type checking $[[v]]$ against $[[uQ]]$ is replaced by 
    a more restrictive premise---checking that $[[v]]$ \emph{synthesizes a subtype} of $[[uQ]]$:
    $$ \ottdruleATArrowAppBidirSLabeled{} $$\\

>>>>

In our system, all type applications are inferred implicitly: the algorithm
\emph{Explicit} type application can be added to the declarative system by the following rule:

$$\ottdruleDTTypeAppLabeled{}$$

@@

\vspace{\baselineskip}
\paragraph*{Solution 1: Declarative Normalization}
One way to resolve this ambiguity is to fix the order of quantifiers.  
The algorithm already performs the ordering of the quantifiers in the normalization procedure.
This way, we could require the inferred type to be normalized
to specify the order of the quantifiers:
$$\ottdruleDTTypeAppOrderedLabeled{}$$

@@

\vspace{\baselineskip}
\paragraph*{Solution 2: Elementary Type Inference}
An alternative approach to explicit type application
 was proposed by \cite{zhao22:elementary}.
In this work, the subtyping relation is restricted 
in such a way that 
$[[∀α⁺.∀β⁺.iN]]$ and $[[∀β⁺.∀α⁺.iN]]$ are \emph{not} mutual subtypes 
(as long as $[[α⁺ ∊ fv(iN)]]$ and $[[β⁺ ∊ fv(iN)]]$).
It implies that the order of the quantifiers of the inferred type is unique,
and thus, the explicit type application is unambiguous.

@@

Elementary type inference
is much more restrictive than what we present.
As mentioned, it forbids the quantifier reordering; besides, 
as soon as the right-hand side of the subtyping relation is
polymorphic, it restricts the instantiation of the left-hand side quantifiers, 
which \emph{disallows}, for example, $[[· ⊢ ∀α⁺.↑α⁺ ≤ ∀α⁺.↑↓↑α⁺]]$.
On the other hand, the elementary subtyping system
can be smoothly extended with bounded quantification, which we discuss later.

\vspace{\baselineskip}
\paragraph*{Solution 3: labeled quantifiers}

A compromise solution that resolves the ambiguity of explicit instantiation
without fixing the order of quantifiers is by using \emph{labeled} quantifiers.
Let us discuss how to introduce them to the \emph{negative} part of the system
since the positive subsystem is symmetric.

@@

The subtyping rule permits an arbitrary order of the quantifiers.
However, similarly to the elementary subtyping, 
each of the right-hand side quantifiers must have 
a matching left-hand side quantifier with the same label. These
matching quantifiers are synchronously removed
(in other words, each variable is abstractly instantiated to itself), 
and the remaining quantifiers are instantiated as usual by
a substitution $[[σ]]$.
$$ \ottdruleDOneLblForallLRLabeled{} $$

@@

Second, notice that the positive subtyping is still not dependent on the negative
subtyping. It means that any leaf-to-root path in the inference tree is monotone:
in the first phase, the negative subtyping rules are applied, and in the second phase, 
only the positive ones.

@@

We do not present the details of the constraint resolution procedure or how it
is integrated into the subtyping algorithm, leaving it for future work. However,
we believe that the system consisting of the mentioned kinds of constraints is solvable
for the following reasons.
\begin{enumerate*}
    \item[(i)] The equivalence entries $[[pua :≈ uP]]$ and $[[nua :≈ uN]]$ are resolved by standard first-order pattern unification techniques.
    \item[(ii)] The resolution of supertyping entries ($[[pua :≥ iP]]$) is discussed in the original algorithm (\cref{sec:constraint-merge}).
    \item[(iii)] The new kind of entries $[[pua :≤ uP]]$ can be
        reduced by case analysis. If $[[uP]]$ is a variable $[[β⁺]]$
        or a shifted computation $[[↓uN]]$ then 
        $[[pua :≤ uP]]$ is equivalent to $[[pua :≈ uP]]$;
        otherwise, $[[pua :≤ ∃nas.↓uN]]$ is equivalently replaced by $[[pua :≈ [nuas/nas]↓uN]]$
        for freshly created $[[nuas]]$.
\end{enumerate*}

@@

\vspace{\baselineskip}
\paragraph*{The Algorithm}
    The algorithmic system is bidirectionalized in a similar way. 
    Each typing premise and the conclusion of the rule are oriented
    to either `checking` ($[[⇐]]$) or `synthesizing` ($[[⇒]]$) mode,
    with respect to the declarative system.

@@

    % \paragraph*{Restricting Inference}
    Instead, we strengthen
    \ruleref{\ottdruleATArrowAppBidirLabel}
    in such a way that it never checks a term against an algorithmic type. Type
    checking $[[v]]$ against $[[uQ]]$ is replaced by a more restrictive
    premise---checking that $[[v]]$ \emph{synthesizes a subtype} of $[[uQ]]$:
    $$ \ottdruleATArrowAppBidirSLabeled{} $$\\
